---
title: "Introduction and Exploratory Data Analysis"
author: "Group 1-Josh Yamamoto, Riley Leonard, Andy Zhao"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(tidymodels)
library(rvest)
library(baseballr)
library(readr)
library(knitr)
```

#Introduction#

In this project, we aim to predict both future salary and earned run average (ERA) of a MLB pitcher using statistics from their past year. The rationale behind this exploration is that, we believe, baseball general managers can become enamored with statistics that are ostensibly important but do not actually predict future performance. For example, a pitcher's total wins is a flashy statistic but the outcome of the game is dependent on several factors outside of the pitcher's control such as their team's offense as well as the opposing pitcher's performance. Thus, the ultimate goal of this project is to investigate if baseball general managers are valuing and paying for the "correct" pitching statistics or the statistics that are actually correlated with or predictive of future performance. To do so, we will fit two models on the same group of predictors, one using salary in year T+1 as a response and the other using ERA in year T+1. We will then compare the relative weights of the various predictors across the two models to investigate any discrepancies.


#Salary Data Exploration#

For data exploration, we will explore the correlations and associations between pitching statistics from the 2016-2019 seasons in order to determine if performance in the past year can predict future salary.

A filter was applied to salary to remove salaries lower than 600,000 or pre-arbitration salaries. These players are under team control and their contracts are artificially deflated. Because their low salaries do not reflect their true "market price", they are excluded.

```{r}
Mega_Summary <- read_csv("Models/Mega Summary.csv")

#Filter out Pre-arbitration
Mega_Summary_Filtered <- 
  Mega_Summary %>% 
  filter(`Salary (t+1)` > 600000)
```
###Looking at Salary###
We began by examining the structure of our first response variable- future salary.

```{r}
ggplot(Mega_Summary_Filtered, 
       aes(x = `Salary (t+1)` )) + 
  geom_histogram(
    bins = 30, 
    alpha = .75 
    ) + 
  theme_bw()
```
As we can see, future salary is quite right skewed. This is understandable as certain superstar pitchers have salaries that are magnitudes greater than the average starting pitcher. We considered a log tranformation to try to ameliorate this skew.

```{r}
ggplot(Mega_Summary_Filtered, 
       aes(x = log(`Salary (t+1)` ))) + 
  geom_histogram(
    bins = 30,
    alpha = .75 ) + 
  theme_bw()
```
It seems like the log transformation greatly reduced the degree of right skew previously observed. While a slight left skew might now exist, we will consider logging salary in our models in order to circumvent this extreme right skew.

###Basic Pitching Statistics###

Wins and losses constitute the most basic of pitching statistics. Total number of pitches is pretty self explanatory too. Additionally, raw, unadjusted ERA, or the average number of runs a pitcher allows in a game is another predictor that one would expect to be associated with salary. Lastly, the number of earned runs, in a similar vein as ERA, is another predictor that stands out as potentially correlated.

```{r}
#Total Wins
ggplot(Mega_Summary_Filtered,
       aes(
         x = W,
         y = `Salary (t+1)`)
       ) +
  geom_point(alpha = 0.5)+
  geom_smooth(method = lm,
              color = "grey50") +
  theme_classic()

cor(Mega_Summary_Filtered$W,
   Mega_Summary_Filtered$`Salary (t+1)`)

#Total Losses
ggplot(Mega_Summary_Filtered,
       aes(
         x = L,
         y = `Salary (t+1)`)
       ) +
  geom_point(alpha = 0.5)+
  geom_smooth(method = lm,
              color = "grey50") +
  theme_classic()

cor(Mega_Summary_Filtered$L,
   Mega_Summary_Filtered$`Salary (t+1)`)

#Total Pitches
ggplot(Mega_Summary_Filtered, 
       aes(
         x = Pitches,
         y = `Salary (t+1)`)
       ) +
  geom_point(alpha = 0.5)+
  geom_smooth(method = lm,
              color = "grey50") +
  theme_classic()

cor(Mega_Summary_Filtered$Pitches,
   Mega_Summary_Filtered$`Salary (t+1)`)

```

Unsurprisingly, wins stand out with a moderate, positive correlation with future salary while losses has, essentially, no relationship with future salary. This pattern makes sense as a pitcher's total number of wins is a seemingly-impressive statistic that many baseball fans would know. Despite the previously-mentioned factors beyond a pitcher's control, GMs are likely to accure high praise by signing a "winning" pitcher, regardless of how much the pitcher contributed to the team's total wins.

Total number of pitches has a moderate, positive correlation with salary. As we will see later on, pitching statistics having to do with productivity or volume of pitches are often quite highly correlated with salary.

```{r}
#Win percent
Mega_Summary_Filtered <- 
  Mega_Summary_Filtered %>% 
  mutate(Win_Percent = W/(W+L)) %>% 
  drop_na()

ggplot(Mega_Summary_Filtered, 
       aes(
         x = Win_Percent,
         y = `Salary (t+1)`)
       ) +
  geom_point(alpha = 0.5)+
  geom_smooth(method = lm, 
              color = "grey50") +
  theme_classic()

cor(Mega_Summary_Filtered$Win_Percent,
   Mega_Summary_Filtered$`Salary (t+1)`)

```

Surprisingly, a pitcher's win percent has a lower correlation with future salary than the total number of wins. While wins and losses are inherently flawed statistics, one would have imagined that a higher win loss ratio suggests a pitcher that deserves to be paid more. Rather, it seems like the total number of wins is the more important statistic in the eyes of GMs.

```{r}
#Games played
Mega_Summary_Filtered2 <-
  Mega_Summary_Filtered %>% 
  filter(G<35)

ggplot(Mega_Summary_Filtered2, 
       aes(
         x = G,
         y = `Salary (t+1)`)
       ) +
  geom_point(alpha = 0.5)+
  geom_smooth(method = lm,
              color = "grey50") +
  theme_classic()

cor(Mega_Summary_Filtered2$G,
   Mega_Summary_Filtered2$`Salary (t+1)`)

```
FIX GAMES AND COME BACK


```{r}
#Current ERA
ggplot(Mega_Summary_Filtered,
       aes(
         x = `ERA`,
         y = `Salary (t+1)`)
       ) +
  geom_point(alpha = 0.5)+
  geom_smooth(method = lm,
              color = "grey50") +
  theme_classic()

cor(Mega_Summary_Filtered$ERA,
    Mega_Summary_Filtered$`Salary (t+1)`)

```

While ERA does have a negative correlation with future salary, the weakness of the relationship is surprising as giving up less runs is an unequivocally positive outcome. The comparison of the relationship between future salary and ERA and future salary and total wins is fascinating as total wins appears to be more highly valued in the eyes of a GM despite a pitcher only having a limited amount of control over their team performance. In contrast, a pichter's ERA is independent of his team's performance, but it has a weaker relationship than total wins. What this suggests is that GMs generally do not do good enough of a job in isolating an individual pitcher's abilities and places too much weight on statistics such as wins which are more stochastic and more susceptible to noise.

###Season Totals and Percentages###

While season totals might have a murkey relationship to future performance, it is undeniable that season totals such as strikeouts can be quite flashy and can lead to lucrative contracts thus might be strong predictors of next year's salary. We first examined total batters faced, total strikeouts, total number of batters walked on balls, total hits, and total home runs given up.

```{r}
#Total batters faced
ggplot(Mega_Summary_Filtered,
       aes(
         x = ABs,
         y = `Salary (t+1)`)
       ) +
  geom_point(alpha=0.5)+
  geom_smooth(method = lm,
              color = "grey50") +
  theme_classic()

cor(Mega_Summary_Filtered$ABs,
    Mega_Summary_Filtered$`Salary (t+1)`)

#Total strikeouts
ggplot(Mega_Summary_Filtered,
       aes(
         x = SO,
         y = `Salary (t+1)`)
       ) +
  geom_point(alpha=0.5)+
  geom_smooth(method = lm,
              color = "grey50") +
  theme_classic()

cor(Mega_Summary_Filtered$SO,
    Mega_Summary_Filtered$`Salary (t+1)`)

#Total batters walked
ggplot(Mega_Summary_Filtered,
       aes(
         x = BB,
         y = `Salary (t+1)`)
       ) +
  geom_point(alpha = 0.5)+
  geom_smooth(method = lm,
              color = "grey50") +
  theme_classic()

cor(Mega_Summary_Filtered$BB,
    Mega_Summary_Filtered$`Salary (t+1)`)

#Total hits
ggplot(Mega_Summary_Filtered,
       aes(
         x = H,
         y = `Salary (t+1)`)
       ) +
  geom_point(alpha = 0.5)+
  geom_smooth(method = lm,
              color = "grey50") +
  theme_classic()

cor(Mega_Summary_Filtered$H,
    Mega_Summary_Filtered$`Salary (t+1)`)

#Total Home Runs
ggplot(Mega_Summary_Filtered,
       aes(
         x = HR,
         y = `Salary (t+1)`)
       ) +
  geom_point(alpha = 0.5)+
  geom_smooth(method = lm,
              color = "grey50") +
  theme_classic()

cor(Mega_Summary_Filtered$HR,
    Mega_Summary_Filtered$`Salary (t+1)`)
```

Unsurprisingly, season totals in general have a strong, positive correlation with salary, and, following a trend that we will see again in this exploration, pitching volume can often be more predictive of future salary than pitch quality. 

Total strikeouts and total batters faced have the strongest positive relationship with future salary. This is to be expected as strikeouts are both the sign of a good pitcher and quite attractive to the average baseball fan and a high number of batters faced suggests a well-regarded pitcher that is entrusted with such responsibilities. More surprisingly, undesirable outcomes like hits and home runs display a medium positive relationship as well. This seemingly contradictory trend perhaps can be explained by the fact that only good pitchers are allowed to rack up hits and home runs as less-accomplished pitchers would simply be cut from the team or sent to the minor league system. If this hypothesis is true, then the batting average against (BAa) should be negatively correlated as BAa is a measurement of what percentage of hitters record a hit against a certain pitcher.

Total number of batters walked on balls has a very weak positive relationship to future salary that is close to 0. This is understandable as number of batters walked can be more contingent upon an individual's style of pitching rather than their quality as a pitcher. Less batters walked simply suggests better control, and superior control in a vacuum does not necessarily suggest a better pitcher.

Next we looked to see if the percent of strikeouts, balls, and hits (BAa) are correlated with future salary

```{r}
#Strikeout Percent
ggplot(Mega_Summary_Filtered,
       aes(
         x = `K %`,
         y = `Salary (t+1)`)
       ) +
  geom_point(alpha = 0.5)+
  geom_smooth(method = lm,
              color = "grey50") +
  theme_classic()

cor(Mega_Summary_Filtered$`K %`,
    Mega_Summary_Filtered$`Salary (t+1)`)

#Percent of Batters walked
ggplot(Mega_Summary_Filtered,
       aes(
         x = `BB %`,
         y = `Salary (t+1)`)
       ) +
  geom_point(alpha = 0.5)+
  geom_smooth(method = lm,
              color = "grey50") +
  theme_classic()

cor(Mega_Summary_Filtered$`BB %`,
    Mega_Summary_Filtered$`Salary (t+1)`)

#Batting average against (hit percent)
ggplot(Mega_Summary_Filtered,
       aes(
         x = BA,
         y = `Salary (t+1)`)
       ) +
  geom_point(alpha = 0.5)+
  geom_smooth(method = lm,
              color = "grey50") +
  theme_classic()

cor(Mega_Summary_Filtered$BA,
    Mega_Summary_Filtered$`Salary (t+1)`)


```

Interestingly, even though strikeout percentage is still positively correlated with future salary, the strength of the relationship is not as strong as the raw number of strikeouts. Both percent of batters walked and hit percent have a moderate-weak negative relationship with future salary. This supports the previous hypothesis that only good, highly-valued pitchers are allowed to accumulate a large number of hits as simply giving up a lot of hits, for example having a high BA, is negatively correlated with future earning. Surprisingly, percent walked has a stronger negative relationship than BA despite total number of hits having a stronger correlation with future salary than total number of batters walked. The discrepancy between total batters walked (BB) and percentage of batters walked (BB%) is not easily explained.

We next looked at some season percentages of more advanced statistics such as hard hit % and barrel %. Both of these statistics look at the speed and angle of the ball after being hit by the batter. Barrel % represents batted balls with exit velocities and launch angles that, historically, have led to a minimum .500 batting average or a 50% chance of resulting in a hit. Hard hit % is similar to barrel % but does not consider launch angle and simply describes batted balls with exit velocities exceeding 95 MPH. Naively, one would expect a negative relationship as a good pitcher should try to avoid giving up hard contact.

```{r}
#Hard hit percent
ggplot(Mega_Summary_Filtered,
       aes(
         x = `Hard Hit %`,
         y = `Salary (t+1)`)
       ) +
  geom_point(alpha = 0.5)+
  geom_smooth(method = lm,
              color = "grey50") +
  theme_classic()

cor(Mega_Summary_Filtered$`Hard Hit %`,
    Mega_Summary_Filtered$`Salary (t+1)`)

#Barrel percent
ggplot(Mega_Summary_Filtered,
       aes(
         x = `Barrel %`,
         y = `Salary (t+1)`)
       ) +
  geom_point(alpha = 0.5)+
  geom_smooth(method = lm,
              color = "grey50") +
  theme_classic()

cor(Mega_Summary_Filtered$`Barrel %`,
    Mega_Summary_Filtered$`Salary (t+1)`)
```

Surprisingly, neither of them seem to have a clear relationship with future salary, barrel %'s relationship is a bit stronger than hard hit % and is, shockingly, positive. However, this weak relationship could be just due to noise as it is unlikely that giving up hard hits will lead to a larger contract.

###Pitch Averages###

Next, we examined if the averages of certain statistics were correlated with salary, specifically looking at average speed and spin rate.

```{r}
#Average Speed
ggplot(Mega_Summary_Filtered,
       aes(
         x = Velocity,
         y = `Salary (t+1)`)
       ) +
  geom_point(alpha = 0.5)+
  geom_smooth(method = lm,
              color = "grey50") +
  theme_classic()

cor(Mega_Summary_Filtered$Velocity,
    Mega_Summary_Filtered$`Salary (t+1)`)

#Average spin rate
ggplot(Mega_Summary_Filtered,
       aes(
         x = `Spin Rate`,
         y = `Salary (t+1)`)
       ) +
  geom_point(alpha = 0.5)+
  geom_smooth(method = lm,
              color = "grey50") +
  theme_classic()

cor(Mega_Summary_Filtered$`Spin Rate`,
    Mega_Summary_Filtered$`Salary (t+1)`)

```

Naively, one might expect faster pitches to be harder to hit and thus expect pitchers with higher average velocities to be handsomely paid. However, the negative correlation, with a not-insignificant correlation, suggests that average pitch speed is actually inversely related to salary. There are a couple of reasons why this might be. The most obvious is that pitchers who lack speed often make up for it with stellar control or a wide variety of possible pitches. Another reason is that, just instinctively, average pitch speed might be correlated with certain negative predictors like more home runs as batters can often make hard contact against fast but poorly-placed pitches.

Spin rate is a topic that has garnered increased attention in the baseball community and is something that pitchers place a lot of emphasis on. It seems that this interest is not misplaced as there is a moderate positive relationship between future salary and spin rate.


###Advanced Statistics###

We now get to our advanced statistics, namely weighted on base average against (wOBA), batting average on balls in play against (BABIP), expected weighted on base average against (xwOBA), and expected batting average against (xBA). The two expected statistics are notable as they try to remove some of the noise or luck from the statistic by calculating WOBA and BA based on the historical WOBA and BA's of batted balls with similar exit velocity and angles. 

```{r}
#Weighted on base average (wOBA)
ggplot(Mega_Summary_Filtered,
       aes(
         x = wOBA,
         y = `Salary (t+1)`)
       ) +
  geom_point(alpha = 0.5)+
  geom_smooth(method = lm,
              color = "grey50") +
  theme_classic()

cor(Mega_Summary_Filtered$wOBA,
   Mega_Summary_Filtered$`Salary (t+1)`)

#Batting average on balls in play against (BABIP)
ggplot(Mega_Summary_Filtered,
       aes(
         x = BABIP,
         y = `Salary (t+1)`)
       ) +
  geom_point(alpha = 0.5)+
  geom_smooth(method = lm,
              color = "grey50") +
  theme_classic()

cor(Mega_Summary_Filtered$BABIP,
   Mega_Summary_Filtered$`Salary (t+1)`)

#Expected wOBA
ggplot(Mega_Summary_Filtered,
       aes(
         x = xwOBA,
         y = `Salary (t+1)`)
       ) +
  geom_point(alpha = 0.5)+
  geom_smooth(method = lm,
              color = "grey50") +
  theme_classic()

cor(Mega_Summary_Filtered$xwOBA,
   Mega_Summary_Filtered$`Salary (t+1)`)

#Expected batting average
ggplot(Mega_Summary_Filtered,
       aes(
         x = xBA,
         y = `Salary (t+1)`)
       ) +
  geom_point(alpha = 0.5)+
  geom_smooth(method = lm,
              color = "grey50") +
  theme_classic()

cor(Mega_Summary_Filtered$xBA,
   Mega_Summary_Filtered$`Salary (t+1)`)

```

Surprisingly, all 4 advanced statistics had negative correlations, albeit weak ones. This suggests that GMs perhaps are not as adverse to advanced statistics as one would expect. The negative relationship is understandable as these statistics measure the offensive production against the pitcher, thus, smaller values and less potent offenses are desirable. Both wOBA and BABIP have stronger relationships with future salary than expected wOBA and expected BA, which suggests that GMs are still not adequately seperating out noise or luck. 

##Luck Reversion##
These predictors will come in handy later but the idea behind luck reversion was that if a pitcher's wOBA was significantly lower than their expected wOBA, then they would appear to be a better pitcher due to luck alone. Luck reversion for us takes 4 forms, the first two (xwOBA-wOBA and xBA-BA) are simply the difference between expected and observed statistics. The second two (ERA/Hard hit% and ERA/barrel %) operate under the assumption that hard contacts (hard hit% and barrel%) will result in more earned runs and a higher ERA, and if a pitcher's ERA/barrel % is low, then that means that they were lucky in terms of having a lower ERA than their pitching truly warrants.

```{r}
#xwOBA-wOBA
ggplot(Mega_Summary_Filtered,
       aes(
         x = `xwOBA - wOBA`,
         y = `Salary (t+1)`)
       ) +
  geom_point(alpha = 0.5)+
  geom_smooth(method = lm,
              color = "grey50") +
  theme_classic()

cor(Mega_Summary_Filtered$`xwOBA - wOBA`,
   Mega_Summary_Filtered$`Salary (t+1)`)

#xBA-BA
ggplot(Mega_Summary_Filtered,
       aes(
         x = `xBA - BA`,
         y = `Salary (t+1)`)
       ) +
  geom_point(alpha = 0.5)+
  geom_smooth(method = lm,
              color = "grey50") +
  theme_classic()

cor(Mega_Summary_Filtered$`xBA - BA`,
   Mega_Summary_Filtered$`Salary (t+1)`)

#ERA/Hard Hit
ggplot(Mega_Summary_Filtered,
       aes(
         x = `ERA/Hard Hit %`,
         y = `Salary (t+1)`)
       ) +
  geom_point(alpha = 0.5)+
  geom_smooth(method = lm,
              color = "grey50") +
  theme_classic()

cor(Mega_Summary_Filtered$`ERA/Hard Hit %`,
   Mega_Summary_Filtered$`Salary (t+1)`)

#ERA/Barrel
ggplot(Mega_Summary_Filtered,
       aes(
         x = `ERA/Barrel %`,
         y = `Salary (t+1)`)
       ) +
  geom_point(alpha = 0.5)+
  geom_smooth(method = lm,
              color = "grey50") +
  theme_classic()

cor(Mega_Summary_Filtered$`ERA/Barrel %`,
   Mega_Summary_Filtered$`Salary (t+1)`)

```

While xwOBA-wOBA doesn't display any strong relationship, xBA-BA had a moderately strong positive relationship, which suggests that lucky players with lower BA's than expected are getting paid more. This further suggests that GMs really have trouble seperating out luck. This trend is further exemplified with ERA/hard hit or barrel as those with smaller values (luckier pitchers) tend to have higher salaries, as suggested by the positive relationship. This relationship is especially pertinent with ERA/Barrel % which has a sizeable correlation coefficient. Overall, it seems like while GMs are attuned to advanced statistics, they have not yet fully bought into luck-adjusted statistics.

##Correlations Between Predictors##

```{r}
total_quant_pred <- 
  select_if(
    Mega_Summary_Filtered,
    is.numeric
    )

total_quant_pred <- 
  total_quant_pred %>% 
  select(
    -Year,
    -ΔERA,-ΔSalary,
    -`Salary (t+1)`
         )

predictor_correlations <- 
  cor(total_quant_pred)

```

In the interest of saving some space, we will not be printing the entire correlation matrix. However, some salient trends stand out in terms of collinearity.

```{r}
volume_predictors<-Mega_Summary_Filtered %>%select(Pitches,W,L,BFP,H,HR,SO)
volume_predictors_cor<-cor(volume_predictors)
kable(volume_predictors_cor)
```

First of all, the "volume" predictors related to sheer number of pitches and games played are all very highly correlated to each other. Specifically, total pitches, total batters faced, total games, total hits, total strikeouts, total wins, total losses, total home runs all have high R values with each other. These are also some of the stronger predictors in predicting salary. In the interest of avoiding collinearity, we might want to combine these variables.

```{r}
hit_predictors<-Mega_Summary_Filtered %>%select(BA,xBA,`K %`)
hit_predictors_cor<-cor(hit_predictors,Mega_Summary_Filtered$`Spin Rate`)
kable(hit_predictors_cor)
```


Interestingly, spin rate is positively correlated with strike percent and negatively correlated with hit percent. So baseball pundits might be on to something in promoting this statistic as increasing strikes and reducing hits is a positive outcome.

```{r}
barrel_predictors<-Mega_Summary_Filtered %>%select(xBA,wOBA,xwOBA,ERA)
barrel_predictors_cor<-cor(barrel_predictors,Mega_Summary_Filtered$`Barrel %`)
kable(barrel_predictors_cor)
```

Average barrel percentage has a medium positive correlation with our advanced statistics wOBA as well as expected wOBA and expected BA. It is also positively correlated with ERA. This makes sense as harder contact means more runs allowed which means more offense, thus raising the values of the advanced statistics.

```{r}
offense_statistics<-Mega_Summary_Filtered %>%select(`K %`,BA,wOBA,xwOBA)
strikes_and_hits<-Mega_Summary_Filtered %>%select(`K %`,BA)
strikes_hits_statistics_cor<-cor(offense_statistics,strikes_and_hits)
kable(strikes_hits_statistics_cor)
```

Strike and hit %'s were negatively correlated with each other which is reasonable as they are mutually exclusive. Strike % also had high, negative correlations with the wOBA and xWOBA statistics while hit % had a high positive correlation with those two statistics. This is understandable as those statistics try to quantify total offensive output, which consists of accumulating hits while avoiding strikes.

```{r}
advanced_statistics<-Mega_Summary_Filtered %>%select(BABIP,xBA,wOBA,xwOBA)
advanced_statistics_cor<-cor(advanced_statistics,Mega_Summary_Filtered$ERA)
kable(advanced_statistics_cor)
```
Our advanced statistics, BABIP, wOBA, xwOBA, xBA are all quite positively correlated with ERA. Given that the statistics all try to quantify offense and that ERA represents total offensive runs allowed, this relationship is to be expected. 

```{r}
wins_and_lossess<-Mega_Summary_Filtered %>%select(W,L)
wins_and_lossess_cor<-cor(wins_and_lossess,Mega_Summary_Filtered$ERA)
kable(wins_and_lossess_cor)
```

An additional, interesting, observation is that ERA is positively correlated with number of losses with a medium degree of positive correlation , while the inverse relationship between ERA and number of wins is not as strong. This suggests that a poor pitcher with a high ERA can easily lose games while a skilled pitcher with a low ERA cannot win games for his team alone.

Here we have a nice way to visualize some of the correlations that we discussed above. Unfortunately due to the large number of predictors we have, the plot is quite cluttered in some ares.

```{r}
library(corrr)

total_quant_pred %>%
  correlate() %>%
  rearrange() %>%
  shave() %>%
  rplot(shape = 15, colours = c("darkorange", "white","midnightblue")) +
  theme(axis.text.x = element_text(angle = 90))
```



#Predicting Future ERA#

Having explored possible predictors of salary (comparing both simple and complex variables), we move on to our prediction of future Earned Run Average (ERA). In this case, our selected response (ERA) can be thought of as a rough proxy for general pitching performance and outcome. Since the inception of the statistic in the early 1900s, ERA has been the most ubiquitous and cited measure of pitcher effectiveness, as it represents a straightforward calculation of the average amount of runs a pitcher allows over the duration of typical game (with run prevention thought of as the ultimate aim of pitching). Despite ERA being the most functional measure of pitching performance, it is nonetheless subject to a significant deal of random noise between individual pitchers' seasons. The development of advanced pitching analytics therefore may lend itself to more accurate forecasts of future ERA than simple counting statistics (such as wins, strikeouts, and innings pitched). With this in mind, we are motivated to find the statistical model that most accurately predicts ERA in a succesive season given a number of predictors generated from data in the current season. This should allow us to assess and predict player performance not based on actual (occasionally random) outcomes, but rather on an aggregation of predicted outcomes determined by a set of relevant statistical parameters. Ultimately, this model should lend itself to making informed salary decisions by MLB general managers.

```{r message=FALSE, warning=FALSE, paged.print=FALSE}
Mega_Summary <- read_csv("Models/Mega Summary.csv")
pitch_data <- read_csv("Exploratory Data Analysis/2020_summary.csv") %>%
  rename(`Pitch Category` = pitch_category) %>%
  rename(wOBA = wOBAA)
```

```{r}
pitch_data  %>%
  ggplot(aes(
    x = `Average Pitch Speed`,
    y = wOBA)
    ) +
  geom_point(alpha = 0.5) +
  geom_smooth(
    method = lm,
    color = "grey50"
    ) +
  theme_classic()
```

```{r}
pitch_data  %>%
  ggplot(aes(
    x = `Average Pitch Speed`,
    y = wOBA, color =`Pitch Category`)
    ) +
  geom_point(
    size = 2, 
    alpha = 0.5) +
  geom_smooth(method = lm) +
  theme_minimal()
```


The first thing we find in our exploration is that average pitch speed has no meaningful relationship with other measures of pitch performance (such as wOBA), even when the data is almost perfectly seperated by categorical variable level. This is also true of average spin rate, and the results are consistent with different response choices (BA, SLG, etc.). This leads us to conclude that pitch tracking data has no predictive power when it comes to modelling pitcher performance. This helps explain why Statcast uses batted ball variables (such as launch angle and exit velocity) rather than pitch tracking variables as the predictors in the calculation of their expected outcome statistics (such as xwOBA and xBA). Accordingly, we will most likely avoid using average pitch speed and average spin rate in our model and focus on other predictors of future ERA instead.


Ostensibly, the most obvious predictor of forcasted ERA is current ERA, as we can intuitively expect pitchers who performed well in a given season to also perform well in subsequent seasons.


```{r}
Mega_Summary %>%
  ggplot(aes(
    x = ERA,
    y = `ERA (t+1)`)
    ) +
  geom_point(alpha = 0.5) +
  geom_smooth(
    method = lm,
    color = "grey50"
    ) +
  theme_classic()
```

```{r}
cor(Mega_Summary$ERA, Mega_Summary$`ERA (t+1)`)
```


Despite the expected existence of a postive linear association between the two variables, the correlation between ERA and ERA in the subsequent season is far from perfect. This can likely be explained by the aforementioned influence of randomness in pitching outcomes. This leads us to believe that other variables (or a combination therof) may have more predictive power in forecasting ERA than ERA by itself. 

##Future ERA distribution##

```{r}
ggplot(Mega_Summary, aes(x = `ERA (t+1)` ))+geom_histogram(bins = 30, alpha = .75 )+theme_bw()
```
We can see that, unlike future salary, future ERA displays a decently normal distribution. While there is a tiny right tail, the low number of observations in that region make it so that it is unlikely to skew our model. This normal distribution is understandable as the MLB has pitchers with a range of talent, with the majority of them falling around the average. The few outliers in the right tail likely represent injury-ridden seasons or seasons where a pitcher experienced bad luck.

##Correlations Between the Strongest Predictors of Salary and Future ERA##

First, we look at some of the same counting stats that were significant in predicting salary, namely total wins, total pitches, and total strikeouts, which were the three predictors with the highest correlation (quantified through R value) with salary:


```{r}
Mega_Summary %>%
  ggplot(aes(
    x = W,
    y = `ERA (t+1)`)
    ) +
  geom_point(alpha = 0.5) +
  geom_smooth(
    method = lm,
    color = "grey50"
    ) +
  theme_classic()
```

```{r}
cor(Mega_Summary$W, Mega_Summary$`ERA (t+1)`)
```


The relationship between wins and forcasted ERA is somewhat weak but the negative or inverse relationship is what one would have expected as winning pitchers often have lower ERAs. Thus, while the negative correlation coefficient does support the decision to reward wins with a lucrative contract, the weaker association suggests that the relationship might not be as important as one would have expected looking at salary alone.

```{r}
Mega_Summary %>%
  ggplot(aes(
    x = Pitches,
    y = `ERA (t+1)`)
    ) +
  geom_point(alpha = 0.5) +
  geom_smooth(
    method = lm,
    color = "grey50"
    ) +
  theme_classic()
```

```{r}
cor(Mega_Summary$Pitches, Mega_Summary$`ERA (t+1)`)
```


Similar to wins, we again see a weak, negative relationship between number of pitches and ERA in the following year. This negative relationship is perhaps a bit more confusing and needs more explanation than the relationship between wins and ERA. What this relationship suggests is that more productive pitchers will tend to have a better or lower ERA next season. This relationship perhaps offers some validation to the correlations between volume of pitches and salary as it suggests that better pitchers simply pitch more often. However, it is worth mentioning that this relationship is more likely due to the fact that only highly-skilled pitchers will be allowed to rack up a large number of total pitchers. That is to say that, while a high number of pitches suggests a more valuable pitcher, making a less-accomplished pitcher pitch more innings will not necessarily improve their value or reduce their ERA.


```{r}
Mega_Summary %>%
  ggplot(aes(
    x = SO,
    y = `ERA (t+1)`)
    ) +
  geom_point(alpha = 0.5) +
  geom_smooth(
    method = lm,
    color = "grey50"
    ) +
  theme_classic()
```

```{r}
cor(Mega_Summary$SO, Mega_Summary$`ERA (t+1)`)
```


Lastly, we see strikeouts in the previous year shows a decently strong, negative relationship to ERA. Here, the relationship is negative, which is what one would expect as pitchers that record more strikeouts are likely to be at the top of their field and are expected to continue their superb performance in later seasons. The strength of this association is surprising as it is stronger than both games and total pitches.

From this analysis, we see that there is a consistently negative relationship between the top three predictors of salary and the pitcher's ERA the next season. This trend provides some evidence that, perhaps, GMs are doing a better job of properly valuing the right attributes than baseball fans imagine as the three predictors that are the most correlated with salary have a positive relationship to future performance (quantified through a negative relationship with future ERA). This finding is quite surprising as, intuitively, these impressive basic counting statistics should provide little to no indication as to future performance. One potential explanation for this is that these basic statistics are correlated with but, crucially, do not cause better future performance. That is to say that only "good" pitchers will be allowed to face a large number of batters and thus will accumulate a large number of these counting statistics.

Despite the fact that the salary-associated predictors are already quite good at predicting future ERA, we hypothesize that advanced statistics such as xWOBA or xBA are going to be better predictors of future performance than the salary-correlated predictors.

##Correlations between Advanced Statistics and Future ERA##

The advanced statistics we considered were strikeout percent (K%), weighted on base average (wOBA), expected weighted on base average (xwOBA), and batting average on balls in play (BABIP).

```{r}
Mega_Summary %>%
  ggplot(aes(
    x = `K %`,
    y = `ERA (t+1)`)
    ) +
  geom_point(alpha = 0.5) +
  geom_smooth(
    method = lm,
    color = "grey50"
    ) +
  theme_classic()
```

```{r}
cor(Mega_Summary$`K %`, Mega_Summary$`ERA (t+1)`)
```


```{r}
Mega_Summary %>%
  ggplot(aes(
    x = wOBA,
    y = `ERA (t+1)`)
    ) +
  geom_point(alpha = 0.5) +
  geom_smooth(
    method = lm,
    color = "grey50"
    ) +
  theme_classic()
```

```{r}
cor(Mega_Summary$wOBA, Mega_Summary$`ERA (t+1)`)
```


```{r}
Mega_Summary %>%
  ggplot(aes(
    x = xwOBA,
    y = `ERA (t+1)`)
    ) +
  geom_point(alpha = 0.5) +
  geom_smooth(
    method = lm,
    color = "grey50"
    ) +
  theme_classic()
```

```{r}
cor(Mega_Summary$xwOBA, Mega_Summary$`ERA (t+1)`)
```

```{r}
Mega_Summary %>%
  ggplot(aes(
    x = BABIP,
    y = `ERA (t+1)`)
    ) +
  geom_point(alpha = 0.5) +
  geom_smooth(
    method = lm,
    color = "grey50"
    ) +
  theme_classic()
```

```{r}
cor(Mega_Summary$BABIP, Mega_Summary$`ERA (t+1)`)
```

Already, we see that the more advanced analytics have, on average, stronger correlations with future ERA than the more rudimentary counting statistics. Additionally, we see that expected weighted on-base average (xwOBA) outperforms weighted on-base average (wOBA). Unlike wOBA, xwOBA attempts to mitigate some of the stochasticity in pitching outcomes by aggregating batted ball data to predict results. In some ways, we can think of xwOBA has a more true measure of pure pitching performance than wOBA, which helps explain why it has a stronger correlation with future ERA. This also allows us to explore the difference between outcome-based statistics (like wOBA) and estimates of expected outcome (like xwOBA) as a measure of a pitcher's "luck", that is how much better or worse batters performed against the pitcher than expected (based on the quality of aontact allowed). 

##Luck Reversion##

Considering that we previously had discussed how total strikeouts is a very potent predictor of future salary, it is unsurprising that strikeout % has the strongest correlation, even stronger than total strikeouts, as it offers a more nuanced approach than the unadjusted total.

A further extension of comparing the differences in outcome and expected outcome is to determine if those differences tend to correct themselves over time. To do so, we analyze the change in a pitcher's ERA from one season to the next as explained by the difference between different predictive and raw measures: 


```{r}
Mega_Summary %>%
  ggplot(aes(
    x = `BABIP - Mean BABIP`,
    y = `ΔERA`)
    ) +
  geom_point(alpha = 0.5) +
  geom_smooth(
    method = lm,
    color = "grey50"
    ) +
  theme_classic()
```

```{r}
cor(Mega_Summary$`BABIP - Mean BABIP`, Mega_Summary$`ΔERA`)
```


```{r}
Mega_Summary %>%
  ggplot(aes(
    x = `xBA - BA`,
    y = `ΔERA`)
    ) +
  geom_point(alpha = 0.5) +
  geom_smooth(
    method = lm,
    color = "grey50"
    ) +
  theme_classic()
```

```{r}
cor(Mega_Summary$`xBA - BA`, Mega_Summary$`ΔERA`)
```


```{r}
Mega_Summary %>%
  ggplot(aes(
    x = `xwOBA - wOBA`,
    y = `ΔERA`)
    ) +
  geom_point(alpha = 0.5) +
  geom_smooth(
    method = lm,
    color = "grey50"
    ) +
  theme_classic()
```

```{r}
cor(Mega_Summary$`xwOBA - wOBA`, Mega_Summary$`ΔERA`)
```


```{r}
Mega_Summary %>%
  ggplot(aes(
    x = `ERA/Hard Hit %`,
    y = `ΔERA`)
    ) +
  geom_point(alpha = 0.5) +
  geom_smooth(
    method = lm, 
    color = "grey50"
    ) +
  theme_classic()
```

```{r}
cor(Mega_Summary$`ERA/Hard Hit %`, Mega_Summary$`ΔERA`)
```


As we see in the data, there is clear statistical evidence of what we could call "mean-reversion of luck". Pitchers whose actual performances were worse than expected performances generally saw an increase in ERA the next season as their "luck" reverted closer to the mean. The opposite can be said for pitchers who performed better than exopcted. 

```{r}
cor(Mega_Summary$`ERA (t+1)`, Mega_Summary$`ΔERA`)
```

Considering the very strong, positive relationship between change in ERA (ΔERA) and future ERA (ERA (t+1)), this conclusion suggests that including differences in outcome-based statistics and expected statistics (especially in accordance with base year ERA) should have significant predictive power in forecasting future ERA.


#Conclusion#

Ultimately, the top 5 highest correlations for future salary are Pitches, ABs, W, H and HR, while the 5 predictors most associated with future ERA are wOBA, BA, xBA, xwOBA,and K %.

```{r}
salary_pred<-Mega_Summary_Filtered %>%select(Pitches, ABs, W, H, HR)
salary_pred_cor<-cor(salary_pred,Mega_Summary_Filtered$`Salary (t+1)`)
kable(salary_pred_cor)
```

```{r}
future_ERA_pred<-Mega_Summary_Filtered %>%select(wOBA, BA, xBA, xwOBA, `K %`)
future_ERA_pred_cor<-cor(future_ERA_pred,Mega_Summary_Filtered$`ERA (t+1)`)
kable(future_ERA_pred_cor)
```

From our preliminary data exploration, it appears that slight discrepancies exist between the predictors that are associated with salary and the predictors associated with future performance. Specifically, it seems like raw, unadjusted counting stats associated with volume of pitches rather than quality are highly correlated with salary and have moderate correlations with future ERA, suggesting that baseball GMs might be prudent in rewarding those statistics. However, advanced statistics, especially statistics that calculate expected values, are better predictors of future performance, despite not being highly correlated with salary. We also elucidated the phenomenon of a "mean-reversion of luck" which suggests that a pitcher's expected performance, derived from statistics such as xWOBA and xBA rather than just WOBA and BA, can help remove some of the noise associated with luck and stochastic fluctuations. Overall, it appears that the differences in strength of correlation between salary and future performance are significant, our next step will be to create and fine-tune models to predict both response variables to both futher delineate what predictors the models differ on as well as investigate both over- and under-paid or valued pitchers.

