---
title: "Extensions and Discussion"
author: "Group 1-Josh Yamamoto, Riley Leonard, Andy Zhao"
date: "12/7/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
pitchers <- read_csv("Final Mega Summary.csv")

custom_mod <- lm(data = pitchers, 
                  ERA_t1 ~ spin_rate + G + SO + K_percent + 
                    ERA:hard_hit_percent + ERA:barrel_percent + luck_adj_ERA)
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
pitchers_2019 <- pitchers %>%
  filter(Year == 2019) %>%
  arrange(desc(salary_t1)) %>%
  head(50) %>%
  select(Pitcher, Year, W, L, G, ERA, spin_rate, SO, BB, K_percent, 
         hard_hit_percent, barrel_percent, luck_adj_ERA, ERA_t1, salary_t1)

custom_mod <- lm(data = pitchers, 
                  ERA_t1 ~ spin_rate + G + SO + K_percent + 
                    ERA:hard_hit_percent + ERA:barrel_percent + luck_adj_ERA)

custom_pred_2019 <- predict(custom_mod, pitchers_2019, type = "response")

custom_pred <- predict(custom_mod, pitchers, type = "response")

pitchers_2019 <- cbind(pitchers_2019, custom_pred_2019)

pitchers <- cbind(pitchers, custom_pred)

normalize_salary <- function(x) {
    (x - min(pitchers_2019$salary_t1)) / (max(pitchers_2019$salary_t1) - min(pitchers_2019$salary_t1))
}

normalize_ERA <- function(x) {
    (x - min(pitchers_2019$custom_pred)) / (max(pitchers_2019$custom_pred) - min(pitchers_2019$custom_pred))
  }


pitchers_2019 <- pitchers_2019 %>%
  mutate(`Standardized Forecasted Adjusted ERA` = -scale(custom_pred_2019)) %>%
  mutate(`Standardized Salary` = scale(salary_t1)) %>%
  mutate(`Normalized Forecasted Adjusted ERA` = normalize_ERA(custom_pred_2019)) %>%
  mutate(`Compensation` = ((salary_t1)/(10 -custom_pred_2019))/100000)

pitchers_2019 <- pitchers_2019 %>%
  mutate(`Standardized Forecasted Adjusted ERA` = format(round(`Standardized Forecasted Adjusted ERA`, 2), nsmall = 2)) %>%
  mutate(`Standardized Salary` = format(round(`Standardized Salary`, 2), nsmall = 2))

pitchers_2019 <- pitchers_2019 %>%
  mutate(`Standardized Forecasted Adjusted ERA` = as.numeric(`Standardized Forecasted Adjusted ERA`)) %>%
  mutate(`Standardized Salary` = as.numeric(`Standardized Salary`))

pitchers_2019$ERA_type <- ifelse(pitchers_2019$`Standardized Forecasted Adjusted ERA` < 0, "below", "above")

pitchers_2019$salary_type <- ifelse(pitchers_2019$`Standardized Salary` < 0, "below", "above")
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
pitchers_2019_tidy <- pitchers_2019 %>%
  pivot_longer(cols = c(custom_pred_2019, ERA_t1), 
               names_to = "type",
               values_to = "value")

#pitchers_2019_tidy$difference <- ifelse(pitchers_2019_tidy$custom_pred_2019  > #pitchers_2019_tidy$ERA_t1 , "lower", "higher")
```


# Extensions and Discussion

### Predicted ERA vs Actual ERA

```{r echo=FALSE, fig.height=6, fig.width=9, message=FALSE, warning=FALSE, paged.print=FALSE}
ggplot(pitchers,
       aes(x = custom_pred, y = ERA_t1)) +
  geom_point(alpha = 0.5, color = "midnightblue") +
  geom_smooth(method = lm, color = "midnightblue") +
  labs(title =  "Predicted ERA vs. Actual ERA (2016-2020)",
       subtitle = "Modelling Predictions",
       x = "Predicted ERA (t+1)",
       y = "ERA (t+1)") + 
  theme_classic()
```
```{r echo=FALSE}
cor(as.numeric(pitchers$custom_pred), pitchers$ERA_t1)
```


```{r echo=FALSE, fig.height=6, fig.width=9, message=FALSE, warning=FALSE, paged.print=FALSE}
ggplot(pitchers,
       aes(x = ERA, y = ERA_t1)) +
  geom_point(alpha = 0.5, color = "midnightblue") +
  geom_smooth(method = lm, color = "midnightblue") +
  labs(title =  "Predicted ERA vs. Actual ERA (2016-2020)",
       subtitle = "Null Predictions",
       x = "ERA",
       y = "ERA (t+1)") + 
  theme_classic()
```
```{r echo=FALSE}
cor(as.numeric(pitchers$ERA), pitchers$ERA_t1)
```



```{r echo=FALSE, fig.width=10,fig.height=6}
ggplot(pitchers_2019_tidy, aes(x = Pitcher,
                          y = value,
                          label = value)) + 
  geom_point(stat = 'identity',
             aes(color = type), 
             size = 2.5) +
  scale_color_manual(name = "Value", 
                    labels = c("Predicted ERA", "Actual ERA"), 
                    values = c("custom_pred_2019" = "deepskyblue", 
                               "ERA_t1" = "palevioletred2")) +
  scale_x_discrete(guide = guide_axis(angle = 50)) +
  labs(title =  "Predicted ERA vs. Actual ERA (2020)",
       x = "Pitcher",
       y = "ERA Value") + 
  geom_line() +
  theme_minimal()
```




### Most Overcompensated and Undercompensated Pitchers Based on Forecasted ERA (2020)

Using forecasted ERA as a measure of predicted performance, we can determine which pitchers were overpaid or underpaid relative to both their contemporaries and their expected pitching outcome. The following visualizations display the standardized forecasted ERA and standardized salaries for the 50 highest-paid starting pitchers in the year 2020, as well as pitcher compensation relative to forecasted ERA:

```{r, echo=FALSE, fig.width=9,fig.height=7}
ggplot(pitchers_2019, aes(x = reorder(Pitcher, `Standardized Forecasted Adjusted ERA`),
                          y = `Standardized Forecasted Adjusted ERA`,
                          label = `Standardized Forecasted Adjusted ERA`)) + 
  geom_bar(stat = 'identity', aes(fill = ERA_type), width = .5)  +
  scale_fill_manual(name = " ", 
                    labels = c("Above Average", "Below Average"), 
                    values = c("above"="deepskyblue", "below"="palevioletred2")) + 
  labs(title = "Diverging Standardized Forecasted ERA (2020)",
       x = "Pitcher",
       y = "Standardized ERA Forecast (Adjusted)") + 
  coord_flip() +
  theme_minimal()
```

```{r, echo=FALSE, fig.width=9,fig.height=7}
ggplot(pitchers_2019, aes(x = reorder(Pitcher, `Standardized Salary`),
                          y = `Standardized Salary`,
                          label = `Standardized Salary`)) + 
  geom_bar(stat = 'identity', aes(fill = salary_type), width = .5)  +
  scale_fill_manual(name = " ", 
                    labels = c("Above Average", "Below Average"), 
                    values = c("above"="deepskyblue", "below"="palevioletred2")) + 
  labs(title = "Diverging Standardized Salary (2020)",
       x = "Pitcher",
       y = "Standardized Salary") + 
  coord_flip() +
  theme_minimal()
```

```{r echo=FALSE, fig.height=8, fig.width=8, message=FALSE, warning=FALSE, paged.print=FALSE}
ggplot() +
  scale_x_continuous(name = "Standardized Forecasted ERA (adjusted)",
                     breaks = seq(-3, 3, 1),
                     limits =c(-3, 3)) +
  scale_y_continuous(name = "Standardized Salary",
                     breaks = seq(-3, 3, 1),
                     limits =c(-3, 3)) +
  ggtitle("Standardized Forecasted ERA vs. Standardized Salary (2020)") +
  geom_rect(aes(xmin = 0, ymin = 0, xmax = 3, ymax = 3), 
            fill = "deepskyblue", alpha = 0.2) +
  geom_rect(aes(xmin = -3, ymin = -3, xmax = 0, ymax = 0), 
            fill = "palevioletred2", alpha = 0.2) +
  geom_density_2d(data = pitchers_2019, 
             aes(x = `Standardized Forecasted Adjusted ERA`, 
                 y = `Standardized Salary`),
             color = "dodgerblue2", alpha = 0.3) +
  geom_point(data = pitchers_2019, 
             aes(x = `Standardized Forecasted Adjusted ERA`, 
                 y = `Standardized Salary`),
             shape = 21,
             colour = "dodgerblue", 
             fill = "grey100", 
             size = 5) +
  geom_text(data = subset(pitchers_2019, 
                          `Standardized Forecasted Adjusted ERA` > 1 | 
                          `Standardized Salary` > 1),
            aes(x = `Standardized Forecasted Adjusted ERA` , 
                y = `Standardized Salary`, 
                label = Pitcher),
            size = 3,
            alpha = 0.8,
            hjust = 0.3,
            vjust = -1.2) + 
  geom_text(data = subset(pitchers_2019, 
                          `Standardized Forecasted Adjusted ERA` < -2),
            aes(x = `Standardized Forecasted Adjusted ERA` , 
                y = `Standardized Salary`, 
                label = Pitcher),
            size = 3,
            alpha = 0.8,
            hjust = 0.3,
            vjust = -1.3) + 
  theme_minimal()
```


```{r echo=FALSE, fig.width=10,fig.height=6}
ggplot(pitchers_2019, aes(x = reorder(Pitcher, Compensation),
                          y = Compensation,
                          label = Compensation)) + 
  geom_col(aes(fill = Compensation)) +
  scale_fill_continuous(type = "viridis") +
  scale_x_discrete(guide = guide_axis(angle = 50)) +
  labs(title =  "Starting Pitcher Compensation (2020)",
       x = "Pitcher",
       y = "Compensation (Salary/Adjusted Forecasted ERA)") + 
  theme_minimal()
```










